# Chapter 3: Logistic regression

- Worked through the DataCamp and finished:

  DataCamp 4 Logistic regression
  
- Worked on R script for Exercise 3
  
  (12.11.-18.11.2019)
  
**Note:** 1.'(Step N)' is used for reviewers to check my work queicky.

2.Somehow I cannot seperate the code of data wrangling from Chapter3.rmd.


## 3.1 Data wrangling

### Read the data file
(Step 1)

```{r}
getwd()
setwd("G:/C-Open Data Science/0-191030/IODS-project-master")
mat <- read.csv(file = "G:/C-Open Data Science/0-191030/IODS-project-master/student-mat.csv", sep = ";", header = TRUE)
```

```{r}
getwd()
setwd("G:/C-Open Data Science/0-191030/IODS-project-master")
por <- read.csv(file = "G:/C-Open Data Science/0-191030/IODS-project-master/student-por.csv", sep = ";", header = TRUE)
```

*Note: I was unable to upload or create a new file named 'data' in github page. Correspondingly, I did not create the data folder and move the two .csv files in.*

### Explore the structure and dimensions of the data

```{r}
head(mat)
dim(mat)
str(mat)
```

```{r}
head(por)
dim(por)
str(por)
```















## 3.2 

