# Chapter 2 Regression and model validation

(5.11.-11.11.2019)

- **Work through the DataCamp**

    DataCamp 3 Regression and model validation

- **Data wrangling**

    R script of data Wrangling exercise is available here.

    https://github.com/yufanyin/IODS-project/blob/master/create_learning2014.R

    The result was saved as 'learning2014.csv'.

**Note:** '(Step N)' is used for reviewers to check my work queicky.


## 2.1 Reading the data

Read the data file (Step 1a)

```{r}
learning2014 <- 
read.table(file = "G:/C-Open Data Science/0-191030/IODS-project-master/learning2014.txt", stringsAsFactors = TRUE) 
str(learning2014)
```

Describe the dataset briefly
(Step 1b)

The data 'learning2014' consists of 166 observations and 7 variables. It contains students' approaches to learning (deep, stra, surf), scores of attitude, points and some background information (gender, age).

## 2.2 Visualizations with ggplot2
(Step 2)

Correlations of the observations between the different variables are shown below with `ggpairs()` function.

```{r}
library(ggplot2)
library(GGally)
p1 <- ggpairs(learning2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)), upper = list(continuous = wrap("cor", size = 2.8)))
p1
```

http://127.0.0.1:42872/chunk_output/DC8623CAB88836EA/A5B1EE4E/cd2f6s6tn8gxk/000002.png

The plot shows the data distribution of all the 166 observations of each variable. The colours represent different gender. Female students are shown in red and male students in verdant.
In the upper right part presents how one variable correlates to another varibable. We are able to find possible significant regression result.

## 2.3 Simple regression with lm()
(Step 3)

**Simple regression: point ~ gender**
Firstly we choose gender as explanatory variables and fit a regression model in which exam points is the dependent variable. Result: not significant.

```{r}
gender_lm <- lm(points ~ gender, data = learning2014)
summary(gender_lm)
```

**Simple regression: point ~ attitude**

```{r}
gender_lm <- lm(points ~ attitude, data = learning2014)
summary(gender_lm)
```

There is a statistical relationship between point and attitude (p:4.12e-09). Then draw a scatter plot of points versus attitude, fit a linear model and print out the summary.

```{r}
qplot(attitude, points, data = learning2014) + geom_smooth(method = "lm")
```

http://127.0.0.1:42872/chunk_output/DC8623CAB88836EA/A5B1EE4E/c5p8rg484lt8q/000002.png

```{r}
model_point_attitude <- lm(points ~ attitude, data = learning2014)
model_point_attitude
```

## 2.4 Multiple regression
(Step 4)

**Multiple regression model: point ~ attitude + stra**
This multiple regression is to test whether strategic learning has an influence on points. The p-value (0.08) shows that the significance of the influence is not very strong. The multiple R-squared is a bit higher than that in regression model of points ~ attitude. That means a little higher correlations if strategic learning is taken into account.

```{r}
model_p_att_stra <- lm(points ~ attitude + stra, data = learning2014)
model_p_att_stra_s <- summary(model_p_att_stra)
model_p_att_stra_s
```

## 2.5 Diagnostic plots
(Step 5)

The results of diagnostic plots are in accordance with the regression model. 

The residuals is close to the 0 line, which means that the relationship is linear. Not many outliers (35, 56,145) are significantly disturbing the analysis but they raise errors in the dataset.

According to Nomal Q-Q plot, most of the values fall on the straight line. The outliers cause randomness in the dataset.

```{r}
plot(model_p_att_stra, which=c(1,2,5))
```

http://127.0.0.1:42872/chunk_output/DC8623CAB88836EA/A5B1EE4E/c4jdrj7wils0g/000004.png?fixed_size=1

http://127.0.0.1:42872/chunk_output/DC8623CAB88836EA/A5B1EE4E/c4jdrj7wils0g/000005.png?fixed_size=1

http://127.0.0.1:42872/chunk_output/DC8623CAB88836EA/A5B1EE4E/c4jdrj7wils0g/000006.png

