# Chapter 6: Analysis of longitudinal data

(3.12.-9.12.2019)

- **Data wrangling**

    R script of data Wrangling exercise is available here.
     
    https://github.com/yufanyin/IODS-project/blob/master/create_BPRS_rats_l.R
    
- **Reference to the data source**
    
    https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt
    
    https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt

    
## 6.1 Load the wide and long data sets

Load and check the data sets

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# load necessary packages
library(tidyr)
library(dplyr)
library(corrplot)
library(ggplot2)
library(GGally)
library(knitr)
library(kableExtra)
library(ggthemes)
library(stringr)
library(lme4)
```

Load the wide (bprs & rats) and long (bprs_l & rats_l) data sets

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
bprs <- read.table("G:/C-Open Data Science/0-191030/IODS-project-master/BPRS.txt", sep = " ", header = TRUE)
rats <- read.table("G:/C-Open Data Science/0-191030/IODS-project-master/rats.txt", header = TRUE)
bprs_l <- read.table("G:/C-Open Data Science/0-191030/IODS-project-master/BPRS_l.txt")
rats_l <- read.table("G:/C-Open Data Science/0-191030/IODS-project-master/rats_l.txt")
```

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
str(bprs_l) 
str(rats_l)
```

bprs data: factor treatment & subject
rats data: factor ID & Group

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
bprs_l$treatment <- factor(bprs_l$treatment)
bprs_l$subject <- factor(bprs_l$subject)
rats_l$ID <- factor(rats_l$ID)
rats_l$Group <- factor(rats_l$Group)
str(bprs_l)
str(rats_l)
```


The BPRS data set

A total 40 male subjects were measured with 2 treatment groups using BPRS (Brief Psychiatric Rating Scale). One BPRS measurement was conducted before treatment started (week0), and weekly measurements followed up to 8 weeks. The BPRS assesses the levels of 18 symptom constructs to evaluate whether patients have schizophrenia  (Vehkalahti & Everitt, 2019). The levels of symptoms constructs were from 1 (not present) to 7 (extremely severe).

The long BPRS data 'bprs_l' 

They consisted of 360 observations in 5 variables. The variables included:  
"treatment": the psychological treatment of the male subjects (1 & 2)  
"subject": the male individuals identification number
"weeks":  the week of treatment and BPRS evaluation
"bprs": the values of bprs
"week": the week number of the BPRS evaluation.

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
knitr::kable(bprs_l, caption = "The long BPRS data 'bprs_l'") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% 
  scroll_box(height = "200px")
```  


The RATS data set
The data consisted of measurements on how rats grew. Three groups of rats got different diets. The body weight of every animal was measured over a period of 9 weeks. The research question is whether the growth profiles of the different diet groups differ (Vehkalahti & Everitt, 2019). 

The long RATS data 'rats_l' 
The data consisted of 176 observations in 5 variables. The variables included:

"ID": the individual rat's identification number
"Group": the diet group (1, 2 & 3)
"WD": the factorial value of the time (the date when rat' weight was measured)
"Weight": weight of individual rat at the given day
"Time": the time point of the weight measurement (days)

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# rats_ data set
knitr::kable(rats_l, caption = "The long RATS data 'rats_l'") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% 
  scroll_box(height = "200px")
```

## 6.2 Graphical displays and summary measure approach

Analyses of chapter 8 of MABS using the RATS data (rats_l)

### 6.2.1 Exploring the data graphically

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
ggplot(rats_l, aes(x = Time, y = Weight,linetype = ID)) +
  geom_line(color = "darkblue") +
  scale_linetype_manual(values = rep(1:6, times= 3)) + 
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "right") +
  scale_y_continuous(limits = c(min(rats_l$Weight), max(rats_l$Weight))) + 
 ggtitle("The long RATS data 'rats_l'")
```



Rat’s starting weights in group 2 & 3 were higher than in group 1 and the weights in group 2 & 3 increased in quite high rates. The weights in group 1 increased in a smaller rate.



### 6.2.2 Standardizing the dataset

Calculating StdWeight and added as a new variable to the 'rats_l'

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_l <- rats_l %>%
  group_by(Time) %>%
  mutate(StdWeight = ((Weight - mean(Weight))/sd(Weight))) %>%
  ungroup()
str(rats_l)
```

Plot the standardised data

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
ggplot(rats_l, aes(x = Time, y = StdWeight, linetype = ID)) +
  geom_line(color = "darkred") +
  scale_linetype_manual(values = rep(1:6, times=3)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "StdWeight: standardized Weight values") +
  ggtitle("Standardized 'rats_l' Weight values") + 
  theme(panel.grid.major.y = element_line(colour = "darkred"))
```


### 6.2.3 Mean profiles

Add the Time as a categorical variable "Time1"

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_l <- rats_l %>%
  mutate(Time1 = factor(rats_l$Time)) %>%
  ungroup()
str(rats_l)
```

The overview of rats' weights over the whole measurement period with boxplots. Different diet groups were in different colors.

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
ggplot(data = rats_l, aes(x = rats_l$Time1, y = Weight, fill = rats_l$Group)) +
  geom_boxplot() +
  ylab("Weight") + 
  xlab("Measurement time [days]") +
  ggtitle("Rat weights") +
  scale_fill_discrete(name = "Diet group") +
  theme(legend.position = "right") 
```

Calculate mean weight of all rats & the standard error of each diet group and add as a variable

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
n_rats <- rats_l$Time %>% unique() %>% length()
rats_s <- rats_l %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = (sd(Weight)/sqrt(n_rats))) %>%
  ungroup()
str(rats_s)
```

Plot the mean profiles

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
ggplot(rats_s, aes(x = Time, y = mean, linetype = Group,  color = Group, shape = Group)) +
  geom_line(size = 0.6) +
  scale_linetype_manual(name = "Diet group", values = c(1,2,3)) +
  geom_point(size=1.5) +
  scale_shape_manual(name = "Diet group", values = c(16,17,18)) +
  scale_color_manual(name = "Diet group", values = c("darkblue", "darkred", "darkgreen")) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.8) +
  theme(legend.position = "right") +
  scale_y_continuous(name = "Mean (Weight) +/- SE (Weight)") + 
  ggtitle("Mean weight profiles of different diet groups")
```

This graph showed the rat weight development of different diet groups during the measurement period. Group 1 started at a lower level and the weight increased slightly . In Group 2 and group 3, weights started from a higher values and then decreased. The standard errors in group 2 & 3 were higher, which indicated that the weight difference between the individual rats was bigger within the diet group.

### 6.2.4 Finding the outlier

Calculating the mean value of weight per diet group

Create a summary data by Group and ID with mean as the summary variable

```{r, rats5, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_l10s <- rats_l %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight)) %>%
  ungroup()
str(rats_l10s)
```


 

## 6.3 Linear mixed effects models for normal response variables

Analyses of chapter 9 of MABS using the BPRS data (bprs_l)


(unfinished, see later)

