# Chapter 2 Regression and model validation

- Worked through the DataCamp and finished:

  DataCamp 3 Regression and model validation

- Worked on R script for Exercise 2:

  Data wrangling and regression analysis

  (5.11.-11.11.2019)
  
**Note:** 1.'(Step N)' is used for reviewers to check my work queicky.

2. Sometimes I was not sure whether I save the figures and tables so that I did not delete many details.



## 2.1 Data wrangling

### Read the data file
(Step 1)

```{r}
setwd("G:/C-Open Data Science/0-191030/IODS-project-master")
lrn14 <- read.table(file = "JYTOPKYS3-data.txt", sep="\t", header=TRUE)
```

**Note:** *I cannot open the connection if*

*read.table(file = "https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS3-data.txt", sep = "\t", header = TRUE)*
                
### Look at the dimensions of the data
(Step 2a)

```{r}
dim(lrn14)
```

### Look at the structure of the data
(Step 2b)

```{r}
str(lrn14)
```

The data consists of 183 observations in 60 variables. Apart from 'gender', all other variables (Age, Attitude, Points, Aa, ... , ST01, ...) are int.

### Create an analysis dataset with the 7 variables
(Step 3a)
### Scale all combination variables to the original scales by taking the mean
(Step 3b)

```{r}
library(dplyr)
deep_questions <- c("D03", "D11", "D19", "D27", "D07", "D14", "D22", "D30","D06",  "D15", "D23", "D31")
strategic_questions <- c("ST01","ST09","ST17","ST25","ST04","ST12","ST20","ST28")
surface_questions <- c("SU02","SU10","SU18","SU26", "SU05","SU13","SU21","SU29","SU08","SU16","SU24","SU32")
```

```{r}
library(dplyr)
deep_columns <- select(lrn14, one_of(deep_questions))
lrn14$deep <- rowMeans(deep_columns)

strategic_columns <- select(lrn14, one_of(strategic_questions))
lrn14$stra <- rowMeans(strategic_columns)

surface_columns <- select(lrn14, one_of(surface_questions))
lrn14$surf <- rowMeans(surface_columns)
```

```{r}
library(dplyr)
keep_columns <- c("gender","Age","Attitude","deep","stra","surf","Points")
learning2014 <- select(lrn14, one_of(keep_columns))
colnames(learning2014)[2] <- "age"
colnames(learning2014)[3] <- "attitude"
colnames(learning2014)[7] <- "points"
colnames(learning2014)

learning2014$attitude <- learning2014$attitude/10
```

### Exclude observations where the exam points variable is zero
(Step 3c)

```{r}
learning2014 <- filter(learning2014, points > 0)
str(learning2014)
```

The data 'learning2014' consists of 166 observations and 7 variables.

```
 'data.frame':	166 obs. of  7 variables:
 gender  : Factor w/ 2 levels "F","M": 1 2 1 2 2 1 2 1 2 1 ...
 age     : int  53 55 49 53 49 38 50 37 37 42 ...
 attitude: num  3.7 3.1 2.5 3.5 3.7 3.8 3.5 2.9 3.8 2.1 ...
 deep    : num  3.58 2.92 3.5 3.5 3.67 ...
 stra    : num  3.38 2.75 3.62 3.12 3.62 ...
 surf    : num  2.58 3.17 2.25 2.25 2.83 ...
 points  : int  25 12 24 10 22 21 21 31 24 26 ...
```

### Save in the iods project folder
(Step 4)

First the the working directory is set and then I save the dataframe with `write.table()` and the `write.csv()`functions.

Set the working directory to IODS project folder with 'setwd()' functions. (Step 4a)

```{r}
setwd("~/IODS-project")
```

Save the analysis dataset to the "data" folder (Step 4b)

```{r}
write.table(learning2014, file = "G:/C-Open Data Science/0-191030/IODS-project-master/learning2014.txt", sep = "\t", col.names = TRUE, row.names = TRUE)
write.csv(learning2014, file = "G:/C-Open Data Science/0-191030/IODS-project-master/learning2014.csv", row.names = FALSE)
```

Check whether the tables can be read using the 'read.table()' and 'read.csv()' function and show the first 6 observations. (Step 4c) 

```{r}
read.table(file = "G:/C-Open Data Science/0-191030/IODS-project-master/learning2014.txt") %>% head()
read.csv(file = "G:/C-Open Data Science/0-191030/IODS-project-master/learning2014.csv") %>%head()
```  

## 2.2 Performing and interpreting regression analysis

### 2.2.1 Reading the data

Read the data file (Step 1a)

```{r}
learning2014 <- 
read.table(file = "G:/C-Open Data Science/0-191030/IODS-project-master/learning2014.txt", stringsAsFactors = TRUE) 

learning2014 %>% str()
```

Describe the dataset briefly (Step 1b)

The data 'learning2014' consists of 166 observations and 7 variables. It contains students' approaches to learning (deep, stra, surf), scores of attitude, points and some background information (gender, age).

### 2.2.2 Visualizations with ggplot2
(Step 2)

Correlations of the observations between the different variables are shown below with `ggpairs()` function.

```{r}
library(ggplot2)
library(GGally)
p1 <- ggpairs(learning2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)), upper = list(continuous = wrap("cor", size = 2.8)))
p1
```

http://127.0.0.1:42872/chunk_output/DC8623CAB88836EA/A5B1EE4E/cd2f6s6tn8gxk/000002.png

The plot shows the data distribution of all the 166 observations of each variable. The colours represent different gender. Female students are shown in red and male students in verdant.
In the upper right part presents how one variable correlates to another varibable. We are able to find possible significant regression result.

### 2.2.3 Simple regression with lm()
(Step 3)

**Simple regression: point ~ gender**
Firstly we choose gender as explanatory variables and fit a regression model in which exam points is the dependent variable. Result: not significant.

```{r}
gender_lm <- lm(points ~ gender, data = learning2014)
summary(gender_lm)
```

**Simple regression: point ~ attitude**

```{r}
gender_lm <- lm(points ~ attitude, data = learning2014)
summary(gender_lm)
```

There is a statistical relationship between point and attitude (p:4.12e-09). Then draw a scatter plot of points versus attitude, fit a linear model and print out the summary.

```{r}
qplot(attitude, points, data = learning2014) + geom_smooth(method = "lm")
```

http://127.0.0.1:42872/chunk_output/DC8623CAB88836EA/A5B1EE4E/c5p8rg484lt8q/000002.png

```{r}
model_point_attitude <- lm(points ~ attitude, data = learning2014)
model_point_attitude
```

### 2.2.4 Multiple regression
(Step 4)

**Multiple regression model: point ~ attitude + stra**
This multiple regression is to test whether strategic learning has an influence on points. The p-value (0.08) shows that the significance of the influence is not very strong. The multiple R-squared is a bit higher than that in regression model of points ~ attitude. That means a little higher correlations if strategic learning is taken into account.

```{r}
model_p_att_stra <- lm(points ~ attitude + stra, data = learning2014)
model_p_att_stra_s <- summary(model_p_att_stra)
model_p_att_stra_s
```

### 2.2.5 Diagnostic plots
(Step 5)

The results of diagnostic plots are in accordance with the regression model. 

The residuals is close to the 0 line, which means that the relationship is linear. Not many outliers (35, 56,145) are significantly disturbing the analysis but they raise errors in the dataset.

According to Nomal Q-Q plot, most of the values fall on the straight line. The outliers cause randomness in the dataset.

```{r}
plot(model_p_att_stra, which=c(1,2,5))
```

http://127.0.0.1:42872/chunk_output/DC8623CAB88836EA/A5B1EE4E/c4jdrj7wils0g/000004.png?fixed_size=1

http://127.0.0.1:42872/chunk_output/DC8623CAB88836EA/A5B1EE4E/c4jdrj7wils0g/000005.png?fixed_size=1

http://127.0.0.1:42872/chunk_output/DC8623CAB88836EA/A5B1EE4E/c4jdrj7wils0g/000006.png

