# Chapter 6: Analysis of longitudinal data

(3.12.-9.12.2019)

- **Data wrangling**

    R script of data Wrangling exercise is available here.
     
    https://github.com/yufanyin/IODS-project/blob/master/create_BPRS_rats_l.R
    
- **Reference to the data source**
    
    https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt
    
    https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt

    
## 6.1 Load the wide and long data sets

Load and check the data sets

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# load necessary packages
library(tidyr)
library(dplyr)
library(corrplot)
library(ggplot2)
library(GGally)
library(knitr)
library(kableExtra)
library(ggthemes)
library(stringr)
library(lme4)
```

Load the wide (bprs & rats) and long (bprs_l & rats_l) data sets

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
bprs <- read.table("G:/C-Open Data Science/0-191030/IODS-project-master/BPRS.txt", sep = " ", header = TRUE)
rats <- read.table("G:/C-Open Data Science/0-191030/IODS-project-master/rats.txt", header = TRUE)
bprs_l <- read.table("G:/C-Open Data Science/0-191030/IODS-project-master/BPRS_l.txt")
rats_l <- read.table("G:/C-Open Data Science/0-191030/IODS-project-master/rats_l.txt")
```

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
str(bprs_l) 
str(rats_l)
```

bprs data: factor treatment & subject
rats data: factor ID & Group

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
bprs_l$treatment <- factor(bprs_l$treatment)
bprs_l$subject <- factor(bprs_l$subject)
rats_l$ID <- factor(rats_l$ID)
rats_l$Group <- factor(rats_l$Group)
str(bprs_l)
str(rats_l)
```


The BPRS data set

A total 40 male subjects were measured with 2 treatment groups using BPRS (Brief Psychiatric Rating Scale). One BPRS measurement was conducted before treatment started (week0), and weekly measurements followed up to 8 weeks. The BPRS assesses the levels of 18 symptom constructs to evaluate whether patients have schizophrenia  (Vehkalahti & Everitt, 2019). The levels of symptoms constructs were from 1 (not present) to 7 (extremely severe).

The long BPRS data 'bprs_l' 

They consisted of 360 observations in 5 variables. The variables included:  
"treatment": the psychological treatment of the male subjects (1 & 2)  
"subject": the male individuals identification number
"weeks":  the week of treatment and BPRS evaluation
"bprs": the values of bprs
"week": the week number of the BPRS evaluation.

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
knitr::kable(bprs_l, caption = "The long BPRS data 'bprs_l'") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% 
  scroll_box(height = "200px")
```  


The RATS data set
The data consisted of measurements on how rats grew. Three groups of rats got different diets. The body weight of every animal was measured over a period of 9 weeks. The research question is whether the growth profiles of the different diet groups differ (Vehkalahti & Everitt, 2019). 

The long RATS data 'rats_l' 
The data consisted of 176 observations in 5 variables. The variables included:

"ID": the individual rat's identification number
"Group": the diet group (1, 2 & 3)
"WD": the factorial value of the time (the date when rat' weight was measured)
"Weight": weight of individual rat at the given day
"Time": the time point of the weight measurement (days)

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# rats_ data set
knitr::kable(rats_l, caption = "The long RATS data 'rats_l'") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% 
  scroll_box(height = "200px")
```

## 6.2 Graphical displays and summary measure approach

Analyses of chapter 8 of MABS using the RATS data (rats_l)

### 6.2.1 Exploring the data graphically

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
ggplot(rats_l, aes(x = Time, y = Weight,linetype = ID)) +
  geom_line(color = "darkblue") +
  scale_linetype_manual(values = rep(1:6, times= 3)) + 
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "right") +
  scale_y_continuous(limits = c(min(rats_l$Weight), max(rats_l$Weight))) + 
 ggtitle("The long RATS data 'rats_l'")
```



Rat’s starting weights in group 2 & 3 were higher than in group 1 and the weights in group 2 & 3 increased in quite high rates. The weights in group 1 increased in a smaller rate.



### 6.2.2 Standardizing the dataset

Calculating StdWeight and added as a new variable to the 'rats_l'

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_l <- rats_l %>%
  group_by(Time) %>%
  mutate(StdWeight = ((Weight - mean(Weight))/sd(Weight))) %>%
  ungroup()
str(rats_l)
```

Plot the standardised data

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
ggplot(rats_l, aes(x = Time, y = StdWeight, linetype = ID)) +
  geom_line(color = "darkred") +
  scale_linetype_manual(values = rep(1:6, times=3)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "StdWeight: standardized Weight values") +
  ggtitle("Standardized 'rats_l' Weight values") + 
  theme(panel.grid.major.y = element_line(colour = "darkred"))
```


### 6.2.3 Mean profiles

Add the Time as a categorical variable "Time1"

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_l <- rats_l %>%
  mutate(Time1 = factor(rats_l$Time)) %>%
  ungroup()
str(rats_l)
```

The overview of rats' weights over the whole measurement period with boxplots. Different diet groups were in different colors.

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
ggplot(data = rats_l, aes(x = rats_l$Time1, y = Weight, fill = rats_l$Group)) +
  geom_boxplot() +
  ylab("Weight") + 
  xlab("Measurement time [days]") +
  ggtitle("Rat weights") +
  scale_fill_discrete(name = "Diet group") +
  theme(legend.position = "right") 
```

Calculate mean weight of all rats & the standard error of each diet group and add as a variable

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
n_rats <- rats_l$Time %>% unique() %>% length()
rats_s <- rats_l %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = (sd(Weight)/sqrt(n_rats))) %>%
  ungroup()
str(rats_s)
```

Plot the mean profiles

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
ggplot(rats_s, aes(x = Time, y = mean, linetype = Group,  color = Group, shape = Group)) +
  geom_line(size = 0.6) +
  scale_linetype_manual(name = "Diet group", values = c(1,2,3)) +
  geom_point(size=1.5) +
  scale_shape_manual(name = "Diet group", values = c(16,17,18)) +
  scale_color_manual(name = "Diet group", values = c("darkblue", "darkred", "darkgreen")) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.8) +
  theme(legend.position = "right") +
  scale_y_continuous(name = "Mean (Weight) +/- SE (Weight)") + 
  ggtitle("Mean weight profiles of different diet groups")
```

This graph showed the rat weight development of different diet groups during the measurement period. Group 1 started at a lower level and the weight increased slightly . In Group 2 and group 3, weights started from a higher values and then decreased. The standard errors in group 2 & 3 were higher, which indicated that the weight difference between the individual rats was bigger within the diet group.

### 6.2.4 Finding the outlier

Calculating the mean value of weight per diet group

Create a summary data by Group and ID with mean as the summary variable

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_l10s <- rats_l %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight)) %>%
  ungroup()
str(rats_l10s)
```

Boxplot of mean weights of different diet groups excluding the first measurement

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
ggplot(rats_l10s, aes(x = Group, y = mean, fill = Group)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape = 23, size = 3, fill = "black") +
  scale_y_continuous(name = "Mean (Weight) / Time 8-64 [days]") + 
  xlab("Diet group") + 
  scale_fill_discrete(name = "Diet group") +
  ggtitle("Mean weights of diet groups excluding the first measurement")
```

The boxplot showed the outliers and they could disturb further comparisons of the different diet groups.

Removing outliers

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_l10s_o <- rats_l10s %>%
    filter(
      (mean > 250 & Group == 1) | 
      (mean < 550 & Group == 2) |
      (mean > 500 & Group == 3))  %>% 
      ungroup()
str(rats_l10s_o)
```

Boxplot of the mean value without the outlier

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
ggplot(rats_l10s_o, aes(x = Group, y = mean, fill = Group)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape = 23, size = 3, fill = "black") +
  scale_y_continuous(name = "Mean (Weight) / Time 8-64 [days]") +
  xlab("Diet group") + 
  scale_fill_discrete(name = "Diet group") +
  ggtitle("Mean weights of diet groups without the outlier")
```

This time there were no outlier values in the boxplot Mean weights of diet groups during 8 to 64 days.

### 6.2.5 Linear regression model and ANOVA

**Note:** I used ANOVA more frequently than t-test so I skip t-test.

Creating a new data table rats_l10s_b. 
The weight baseline values were created by the first weigh measurement on day 1 (WD1) as a new variable to the summary data. 

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_l10s_b <- rats_l10s %>%
  mutate(baseline = rats$WD1)
rats_l10s_b %>% knitr::kable() %>% kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% scroll_box(height = "200px")
``` 


To find which diet group had the highest increase in weight during the measurement period,  anova tests were done using this new data 'rats_l10s_b' with the mean weight values and the baseline weight values.

Linear regression model of mean values in different diet groups

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_lrm1 <- lm(mean ~ Group, data = rats_l10s_b)
round(rats_lrm1$coefficients, digits = 2) %>% knitr::kable() %>% kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")
```

The model showed that the mean value of group 2 was 222g higher and the mean value of group 3 was 262g higher than that of group 1.

Examine the group means using a post-hoc test after the ANOVA to find whether the diet groups differed to each other.

The model shows that the mean value of group 2 is 222g higher and the mean value of group 3 is 262g higher compared to group 1 mean weights. Let's have a look now if the diet groups differ to each other. With a post-hoc test after the ANOVA we can check the group means.

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_anova1 <- anova(rats_lrm1)
rats_anova1
```

The ANOVA of the linear model (mean weight ~ diet groups) showed that group 2 had a significant different weight increase compared to group 1 & 3.

Perform an ANOVA and a post-hoc test (TukeyHSD) to find whether group means of diet groups were different from each other. The dataset `rats_l10s_b` was without the first measurement (start value) and inluded the outlier values.

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_anova2 <- aov(mean ~ Group, data = rats_l10s_b)
summary(rats_anova2)
TukeyHSD(rats_anova2, "Group", ordered = TRUE, conf.level = 0.95)
```

This post-hoc test provided the differences between the diet groups. Group 1 & 2 were significantly different in mean weight. So were Group 1 & 3. Group 2 and 3 were not significantly different in the mean weight mean, influenced by the outlier.

The difference of each group to the baselines:

The linear model with the mean as the response

```{r, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_lrm2 <- lm(mean ~ baseline + Group, data = rats_l10s_b)
round(rats_lrm2$coefficients, digits = 2) %>% knitr::kable() %>% kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")
```

That was the regression model of mean weights compared to the weight baseline (the first weight measurement of each dietary group). The results shows that the weight difference between baseline and group 2 was 34g. The weight difference between baseline and  group 3 and baseline was 23g. Therefore Group 2 has the highest weight increase rates during the measurement period.

```{r, rats14, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
rats_anova3 <- anova(rats_lrm2)
rats_anova3
```

The baseline had a significant influence in the weight increase in different diet groups. Group 2 had a significant weight increase relative to the baseline, compared to group 1 & 3. Diet 2 may lead to the highest weight increase.

## 6.3 Linear mixed effects models for normal response variables

Analyses of chapter 9 of MABS using the BPRS data (bprs_l)


(unfinished, see later)

